%\newcommand{\myyear}{2004-2015}
% 2004 for all but ISA ... er ... and memory (290 notes)
\newcommand{\myyear}{2000-2022}
\include{ece120-notes-macros}

\newcommand{\twovert}[2]{\mbox{\arraycolsep 0pt \tabcolsep 0pt \begin{tabular}{c}{#1}\\{#2}\end{tabular}}}
\newcommand{\threevert}[3]{\mbox{\arraycolsep 0pt \tabcolsep 0pt \begin{tabular}{c}{#1}\\{#2}\\{#3}\end{tabular}}}
\newcommand{\fourvert}[4]{\mbox{\arraycolsep 0pt \tabcolsep 0pt \begin{tabular}{c}{#1}\\{#2}\\{#3}\\{#4}\end{tabular}}}
\newcommand{\fivevert}[5]{\mbox{\arraycolsep 0pt \tabcolsep 0pt \begin{tabular}{c}{#1}\\{#2}\\{#3}\\{#4}\\{#5}\end{tabular}}}

\begin{document}

\pagenumbering{roman}
\thispagestyle{empty}


\mbox{~~~}\vspace{3in}

\centerline{\Huge \bf ECE~120}
\vspace{20pt}
\centerline{\huge \bf Introduction to Computing}
\vspace{30pt}
\centerline{\LARGE \bf Department of Electrical and Computer Engineering}
\vspace{10pt}
\centerline{\LARGE \bf University of Illinois at Urbana-Champaign}
\vspace{10pt}
\centerline{\Large \bf \copyright~\myyear~Steven S. Lumetta.  All rights reserved.}
\vfill\vfill

\pagebreak

The bulk of these notes was developed for the ECE~120 class at the
University of Illinois at Urbana-Champaign, but some material was taken
from my previous lecture notes for 
ECE~290 (written around 2000),
ECE~190 (written around 2004), and 
ECE~391 (written around 2004).

Doug Jones was my co-conspirator in developing the new design, and both 
he and Geoffrey Herman offered frequent and helpful feedback on early 
drafts, for which I am grateful.  Numerous other instructors have also
given me feedback over the years, including but not limited to 
%
% S13
Volodymyr Kindratenko,
% in between
Wen-mei~W. Hwu,
% F16
Suma Bhat,
Juan Jos\'e Jaramillo,
David Varodayan,
% S17
Donna Brown,
% F20
Ujjal Bhowmik,
and
Kirill Levchenko.
%
Where the notes are more easily read, thanks are due to them.  Remaining
mistakes should be attributed to me.

As part of ECE Illinois' determination to make our curricula more inclusive,
I have also decided to consciously identify and eliminate any non-inclusive
and potentially offensive terminology from our course materials.  In the
rare cases in which I feel that students may encounter such terms in their
careers, I may mention their previous use.  I also encourage you to report
any remaining instances to me directly.

\pagebreak


\pagestyle{headings}
\markboth{\copyright~\myyear~Steven S. Lumetta.  All rights reserved.}{\copyright~\myyear~Steven S. Lumetta.  All rights reserved.}

\tableofcontents
\vfill
\pagebreak
%\mbox{~~~} % blank page
%\pagebreak

%\pagestyle{headings}
\markboth{\copyright~\myyear~Steven S. Lumetta.  All rights reserved.}{\copyright~\myyear~Steven S. Lumetta.  All rights reserved.}

\pagenumbering{arabic}
\setcounter{page}{1}

\setcounter{section}{1}
\include{part1/ece120-set-1-1-halting-problem}
\include{part1/ece120-set-1-2-twos-complement}
\include{part1/ece120-set-1-3-overflow}
\include{part1/ece120-set-1-4-logic}
\include{part1/ece120-set-1-5-programming}
\include{part1/ece120-set-1-6-summary}

\setcounter{section}{2}
\setcounter{subsection}{0}
\include{part2/ece120-set-2-1-goodforms}
\include{part2/ece120-set-2-2-dontcare}
\include{part2/ece120-set-2-3-adder}
\include{part2/ece120-set-2-4-comparator}
\include{part2/ece120-set-2-5-abstraction}
\include{part2/ece120-set-2-6-sequential}
\include{part2/ece120-set-2-7-registers}
\include{part2/ece120-set-2-8-summary}

\setcounter{section}{3}
\setcounter{subsection}{0}
\include{part3/ece120-set-3-1-serialize}
\include{part3/ece120-set-3-2-fsm-examples-part-i}
\include{part3/ece120-set-3-3-lab}
\include{part3/ece120-set-3-4-keyless-extension}
\include{part3/ece120-set-3-5-fsm-examples-part-ii}
\include{part3/ece120-set-3-6-memory}
\include{part3/ece120-set-3-7-fsm-to-computer}
\include{part3/ece120-set-3-8-summary}

\setcounter{section}{4}
\setcounter{subsection}{0}
\include{part4/ece120-set-4-1-control-unit}
%\setcounter{subsection}{1}
%\setcounter{page}{145}
\include{part4/ece120-set-4-2-coding}
\include{part4/ece120-set-4-3-isa-design}
\include{part4/ece120-set-4-4-summary}

\end{document}
